A Stopped Negative Binomial Distribution

### 2016-08-16 stat-and-prob-letters/snb-paper.tex
Improved hypergeometric distribution interpretation.

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. The binomial probability of a patient 
responding to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject this hypothesis and the treatment is deemed successful at 
significance level of $0.1$.  If fewer than seven respond then the null 
hypothesis is not 
rejected and the treatment is deemed ineffective.

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
successful patients is one endpoint and the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is seven through 17.

@@

A hypothetical sample path is illustrated in Fig.~\ref{fig:kane_viz}.
The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients 
enrolled. The horizontal and vertical boundaries represent
endpoints for the trial. In this case a seventh response was reached on
the $15^{\text{th}}$ enrollment.
Since the success boundary is reached, we say the treatment succeeds.

@@

When $p=0$ or $1$ the processes are deterministic and variance is zero.
Value of $p$ between zero and one change the mixing proportions of 
the two endpoints. The saddle around $p=0.25$ results from inequality 
in the size of the support of the two endpoints.

@@

In the rest of this work we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). 
%Some of its characteristics are summarized in Tab.~\ref{tab:snb}.
This paper derives this distribution and explores its properties.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the predictive and posterior distributions when $p$ has a 
beta prior distribution.

@@

The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1967}.
The maximum negative binomial \cite{Zhang2000,Zelterman2005} is the smallest 
number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders and is equivalent
to a translated version of the riff-shuffle.

@@

The SNB's ability to approximate the geometric, normal, gamma, and Poisson
distributions follow from it generalizing the negative binomial distribution. 
\begin{prop}
The MGF of the SNB converges to that of the negative binomial when either
$s$ or $t$ gets large. That is
\begin{equation*}
\mathbb{E} e^{xY} \rightarrow \left( \frac{pe^x}{1-qe^x} \right)^s
\end{equation*}
as $t \rightarrow \infty$. The analogous result holds when 
$s \rightarrow \infty$.
\end{prop}
\begin{proof}
The second regularized incomplete beta term in (\ref{eqn:mgf}) can be written
in terms of a cumulative binomial distribution
\begin{equation*}
\mathcal{I}_{1-pe^x}(t, s) = \mathbb{P}\left[ B \leq s-1 \right]
\end{equation*}
where $B$ is distributed as
Binomial($t-k$, $pe^x$). By Hoeffding's inequality \cite{Hoeffding1963}
it follows that
\begin{equation} \label{eqn:hoeffding}
\mathbb{P}\left[ B \leq s-1 \right] \leq 
  \exp \left(-2\frac{((t-k)pe^x - s-1)^2}{t-k} \right).
\end{equation}
As $t$ gets large $\mathcal{I}_{1-pe^x}(t, s)$ goes to zero
Likewise, $\mathcal{I}_{1-qe^x}(s, t)$ goes
to one. The proof follows by realizing 
\begin{equation*}
0 < \frac{qe^x}{1-pe^x} < 1
\end{equation*}
over the support of $x$.
\end{proof}

@@

Drawing connections to the gamma and Poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Ord1968,Guenther1972,Best1974} for examples). 
A connection to the Poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a Poisson is proportional
to a chi-square distribution with $2k$ degrees of freedom
then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between betwen the cumulant generating
functions converge to zero.

The lower-half normal distribution can be approximated by setting $s=t$
for appropriately large $s$ and $t$ and $p = 0.5$. In this
case the SNB can be viewed as identical, negative binomials
approximating a normal and truncated at the median.

@@

Let us consider the posterior distribution of $p$ with an SNB likelihood 
and Beta($\alpha$, $\beta$) prior distribution.
The prior times the likelihood is given by the function
\begin{align} \label{eqn:ptl}
f_p(x | \alpha, \beta ) f_{Y|p}(x | k, s, t) &= 
  \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} x^{\alpha +s -1} 
    (1-x)^{k+\beta-s-1} + \\
  & \frac{ {k-1 \choose t-1} }{B(\alpha, \beta)} p^{k+\alpha -t -1} 
    (1-x)^{\beta+t-1} \nonumber
\end{align}
where $0 \leq x \leq 1$, $s \leq k \leq s+k-1$, and $t \leq k \leq s+k-1$.
This result can be found directly by multiplying the probability mass function
in (\ref{eqn:pmf}) by the density of the Beta distribution.

The predictive distribution of the SNB can be found by integrating
$x$ over the interval zero to one and applying the definition of the 
beta function.
\begin{align} \label{eqn:predictive}
\int_0^1 f_p(x | \alpha, \beta )  f_{Y|p}(x | k, s, t) & dx = 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    + \nonumber \\
  & {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)}.
\end{align}

The ratio of (\ref{eqn:ptl}) divided by (\ref{eqn:predictive}) gives the 
posterior distribution of $p$. It is a mixture of beta distributions. The
mixing parameter depend on the endpoints ($s$ and $t$), the number of enrollees needed to reach an endpoint ($k$), and the prior parameters ($\alpha$ and
$\beta$).

@@

To better understand the predictive distribution, let us once again return to 
our example where $s=7$, $t=11$. However, we assume $p$ has a beta distribution
with shape parameters $\alpha = 0.2 c$ and $\beta = 0.8 c$ for any $c > 0$.
This parameterization has mean 0.2 for all $c > 0$ and variance 
$0.16 / (c+1)$.

\begin{figure}[h!]
\centering
\subfloat[mean]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[variance]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The mean and variance of the predictive SNB with $s=7$, $t=11$
with $\mathbb{E}\ p = 0.2$ varying the uncertainty 
in the distribution of $p$. A larger value of $c$
corresponds to a beta variance.
}
\label{fig:bayesian-sample-size}
\end{figure}

As $c$ increases the 
predictive distribution sample statistics converge to corresponding 
statistics for the SNB with $p = 0.2$. 
For estimation it may be 
preferable to select shape parameters based on the median or mode of the 
beta distribution since they are less sensitive to distributional uncertainty.

>>>>

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. The binomial probability of a patient 
responding to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject this hypothesis and the treatment is deemed successful at 
a significance level of $0.1$.  If fewer than seven respond then the null 
hypothesis is not 
rejected and the treatment is deemed ineffective.

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
successful patients is one endpoint and the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling, the range of the
sample size for this trial is seven through 17.

@@

A hypothetical sample path is illustrated in Fig.~\ref{fig:kane_viz}.
The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients 
enrolled. The horizontal and vertical boundaries represent
endpoints for the trial. In this case, a seventh response was reached on
the $15^{\text{th}}$ enrollment.
Since the success boundary is reached, we say the treatment succeeds.

@@

When $p=0$ or $1$ the processes are deterministic and variance is zero.
Values between zero and one change the mixing proportions of 
the two endpoints. The saddle around $p=0.25$ results from inequality 
in the size of the support of the two endpoints.

@@

In the rest of this work, we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). 
%Some of its characteristics are summarized in Tab.~\ref{tab:snb}.
This paper derives this distribution and explores its properties.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the predictive and posterior distributions when $p$ has a 
beta prior distribution.

@@

The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial 
distribution~\citep{Uppuluri1967,Johnson2005}.
The maximum negative binomial \cite{Zhang2000,Zelterman2005,Johnson2005} is 
the smallest number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders and is equivalent
to a translated version of the riff-shuffle.

@@

The SNB's ability to approximate the geometric, normal, gamma, and Poisson
distributions follow from it generalizing the negative binomial distribution. 
\begin{prop}
The MGF of the SNB converges to that of the negative binomial when either
$s$ or $t$ gets large. That is
\begin{equation*}
\mathbb{E} e^{xY} \rightarrow \left( \frac{pe^x}{1-qe^x} \right)^s
\end{equation*}
as $t \rightarrow \infty$. The analogous result holds when 
$s \rightarrow \infty$.
\end{prop}
\begin{proof}
The second regularized incomplete beta term in (\ref{eqn:mgf}) can be written
in terms of a cumulative binomial distribution
\begin{equation*}
\mathcal{I}_{1-pe^x}(t, s) = \mathbb{P}\left[ B \leq s-1 \right]
\end{equation*}
where $B$ is distributed as
Binomial($t-k$, $pe^x$). By Chebychev's inequality %\cite{Hoeffding1963}
it follows that
\begin{equation} \label{eqn:hoeffding}
\mathbb{P}\left[ B \leq s-1 \right] \leq 
  \frac{ (t-k) pe^x (1-pe^x) }{ \left(s - (t-k)pe^x\right)^2 }
\end{equation}
As $t$ gets large $\mathcal{I}_{1-pe^x}(t, s)$ goes to zero
Likewise, $\mathcal{I}_{1-qe^x}(s, t)$ goes
to one. The proof follows by realizing 
\begin{equation*}
0 < \frac{qe^x}{1-pe^x} < 1
\end{equation*}
over the support of $x$.
\end{proof}

@@

Drawing connections to the gamma and Poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Ord1968,Guenther1972,Best1974} for examples). 
A connection to the Poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a Poisson is proportional
to a chi-square distribution with $2k$ degrees of freedom
then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between the cumulant generating
functions converge to zero.

The lower-half normal distribution can be approximated by setting $s=t$
for appropriately large $s$ and $t$ and $p = 0.5$. In this
case, the SNB can be viewed as identical, negative binomials
approximating a normal and truncated at the median.

@@

Let us consider the case where the rate parameter is distributed 
as Beta($\alpha$, $\beta$) and denoted $P$.
The prior times the likelihood is given by the function
\begin{align} \label{eqn:ptl}
f_P(p, \alpha, \beta ) f_{Y|P}(p, k, s, t) &= 
  \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1} + \\
  & \frac{ {k-1 \choose t-1} }{B(\alpha, \beta)} p^{k+\alpha -t -1} 
    (1-p)^{\beta+t-1} \nonumber
\end{align}
where $0 \leq p \leq 1$, $s \leq k \leq s+k-1$, and $t \leq k \leq s+k-1$.
This result can be found directly by multiplying the probability mass function
in (\ref{eqn:pmf}) by the density of the Beta distribution.

The predictive distribution of the SNB can be found by integrating
$p$ over the interval zero to one and applying the definition of the 
beta function.
\begin{align} \label{eqn:predictive}
f_{Y}(k, s, &t, \alpha, \beta) = 
  \int_0^1 f_P(p | \alpha, \beta )  f_{Y|P}(p, k, s, t) dp \nonumber \\ 
 & = {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    + {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)}.
\end{align}
If both $\alpha$ and $\beta$ are non-negative integers then the predictive
distribution is a mixture of hypergeometric distributions. 
\begin{align*} \label{eqn:hypergeo}
f_{Y}(k, s, t, \alpha, \beta) = 
  \frac{ {k - 1 \choose s - 1 } 
         {\alpha + \beta \choose \alpha} }{
         {\alpha + \beta + k - 1 \choose \alpha + s}}
  \frac{\alpha}{\alpha + \beta}
  \frac{ \beta }{ k-s+\beta } +
  \frac{ {k - 1 \choose t - 1} 
         {\alpha + \beta \choose \beta} 
         }{
         {\alpha + \beta + k -1 \choose t + \beta}
       }
  \frac{ \beta }{ \alpha + \beta}
  \frac{ \alpha }{ k-t + \alpha}
\end{align*}
The ratio of combinations in the first term can be interpreted as the
probability of $s-1$ responders from $k-1$ patients in $\alpha + s$ draws
from a population size of $\alpha + \beta + k - 1$. This value is multiplied
by $\alpha / (\alpha + \beta)$, the expected response rate of the prior. 
The final term in the product weigths the prior based on the number
of non-responders ($k-s$). Terms in the second summand are interpreted similarly
for non-responders.

The ratio of (\ref{eqn:ptl}) divided by (\ref{eqn:predictive}) gives the 
posterior distribution of $P$. It is a mixture of beta distributions. The
mixing parameter depend on the endpoints ($s$ and $t$), the number of enrollees needed to reach an endpoint ($k$), and the prior parameters ($\alpha$ and
$\beta$).

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{beta-mixture.pdf}
\caption{
The posterior distribution of the response probability with 
the Jeffreys prior ($\alpha=\beta=1/2$) for the trial where
$s=7$, $t=11$, $k=15$.
}
\label{fig:beta_mixture}
\end{figure}

The posterior result above is for the case where the parameters are known
and the endpoint is not. That is, we do not which boundary was reached.
Fig.~\ref{fig:beta_mixture} shows this distribution based on the 
parameters in the hypothetical trial assuming the
Jeffreys prior \citep{Jeffreys1946}.  If we include the fact that the
trajectory reaches the endpoint then the second terms in (\ref{eqn:ptl}) and
(\ref{eqn:predictive}) are both zero and the posterior distribution is
Beta(7.5, 8.5). This is proportional to the area labeled ``success'' in 
Fig.~\ref{fig:beta_mixture}.

@@

%To better understand the predictive distribution, let us once again return to 
%our example where $s=7$, $t=11$. However, we assume $p$ has a beta distribution
%with shape parameters $\alpha = 0.2 c$ and $\beta = 0.8 c$ for any $c > 0$.
%This parameterization has mean 0.2 for all $c > 0$ and variance 
%$0.16 / (c+1)$.
%
%
%\begin{figure}[h!]
%\centering
%\subfloat[mean]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
%\hfill
%\subfloat[variance]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
%\caption{
%The mean and variance of the predictive SNB with $s=7$, $t=11$
%with $\mathbb{E}\ p = 0.2$ varying the uncertainty 
%in the distribution of $p$. A larger value of $c$
%corresponds to a beta variance.
%}
%\label{fig:bayesian-sample-size}
%\end{figure}

%As $c$ increases the 
%predictive distribution sample statistics converge to corresponding 
%statistics for the SNB with $p = 0.2$. 
%For estimation it may be 
%preferable to select shape parameters based on the median or mode of the 
%beta distribution since they are less sensitive to distributional uncertainty.
